Diffusion of Innovation in the Open Source Software Ecosystem

### 2014-08-25 Paper/paper.tex
Quite a few changes 1. Added title.tex - unclear if we need to remove author names from the main paper 2. Added a few more articles to paper.bib - include the Meade and Islam paper 3. Justified our statistical use of NLS (NonLinearFit), we can use a variant of this as suggested by Meade and Islam 4. Removed most typos in the paper.

\title{Diffusion of Innovation in the Open Source Software Ecosystem}
\subtitle{An Empirical Study of the Gentoo Package Database} 
\dedication{}
\author{Remco Bloemen \and Chintan Amrit}
\institute{University of Twente}

@@

In this paper we make the case that software package dependencies are a form of  adoption of innovation. We then test this on the time-evolution of the Gentoo package dependency graph. We find that the Bass model of innovation diffusion fits the growth of the number of packages depending on a given library with an $\overline{R}^2 > 98\%$.
\end{abstract}

@@

In this paper we not only analyse the adoption of open source libraries by other projects, but also use a method to test the diffusion of innovation of a software project - or the lifecycle of a project's adoption. We analyse the adoption of the open source libraries using the theoretical lens of \emph{diffusion of innovation} \citep{rogers03} and test it with a particular formalisation borrowed from the field of marketing research namely the \emph{Bass diffusion model} \citep{bass69}. Our central hypothesis is that  \emph{the growth in the number of open source software dependees of a project (its adoption) is a Bass innovation diffusion process.} In particular, we study  software innovation of the libraries in the Gentoo Portage  database from an adoption perspective. We first download the Gentoo Portage database for a time line of one month over a period of around 13 years (from 1 January to around mid 2012). We then calculate the dependencies among the libraries in each version of the database. Then, we calculate the change in the dependency graph of some well known libraries and check if this adoption of the library fits the Bass diffusion model \citep{bass69}.  We then analyse the adoption of 10 randomly selected libraries (we discard and replace those libraries whose adoption trends contain a lot of noise due to various reasons) compared to the Bass diffusion model. In all cases we find that the Bass diffusion model explains the adoption of the libraries (with an $\overline{R}^2 > 98\%$). This finding shows that the adoption of open source software libraries (in the Gentoo Portage database) does not differ from the adoption of other kinds of innovations as predicted by the theory of adoption of innovation, and in particular the Bass diffusion model. This finding, if found generalisable to most large open source libraries, can help in predicting their adoption over their lifetime. This can hence provide guidance to package managers and users of open source libraries.

% (Adoption reflects innovation 'better' than changes within packages. - Still cannot find a relevant reference!)

In the next section (section 2)  we describe the background literature of the paper, this is followed by the details of our method of extracting the dependencies of libraries in section 3. In section 4 we analyse the dataset and test the fit of the trends of library adoption with the Bass diffusion model. Section 5 discusses the results and in section 6 we conclude the paper by discussing the limitations of this study and provide some guidance for future work.

@@

In \citet{crowston08} published a comprehensive overview of academical research on open source software development. Of the 184 articles they cite, the vast majority of articles are case studies or surveys, with 4\% of the articles describing the development of empirical instruments and/or measurements. Also, most articles look at the level of a particular group or project, with 4\% looking at the societal level of interacting projects. Crowston et al. (2008) find no articles that develop instruments or measurements on the level of OSS packages \citep{crowston08}, which we aim to do in this paper.

@@

In figure~\ref{fig:bass} two Bass diffusions are plotted using equation \eqref{eq:bass-direct}, one representing an innovation diffusion with many innovators and a few imitators and one representing a diffusion with many imitators and a few innovators. When there are more imitators it takes a while for the innovation to be adopted since the majority of the potential users are waiting for someone else to try it first.

@@

The interpretation of the variables and parameters and their dimensions is presented in table~\ref{tbl:bass-model}. When applying the formula, one should note that it is non-linear, so ordinary linear regression can not be used. Instead one can use non-linear least squares regression, while noting the correct number of the degrees of freedom. This can be done using existing mathematical/statistical packages. In this paper we used the \verb|NonlinearModelFit| procedure in Mathematica.

@@

By design the database can work with complicated dependency relations, such as ``package \verb|amarok| requires package \verb|phonon-kde|, minimum version \verb|4.3|, but only when feature \verb|player| is required''. This would be coded as \verb|player?| \verb|(| \verb|>=kde-base/phonon-kde-4.3| \verb|)|. An example of the run time dependencies for the Amarok music player is given in figure~\ref{fig:amarokruntime}. The figure includes more complex rules such as ``either package X or package Y is required'', `` package X is required to contain feature Y'', etcetera.

@@

In figure~\ref{fig:pkgsgrowth} the number of packages in the database is plotted over time. One can see how the database started in 2001, underwent a period of rapid growth between 2002---2006 and settled into calm a linear growth from 2006 onwards. In figure~\ref{fig:pkgsgrowthdelta} the rate of introducing and removing packages is plotted over time. This shows two spikes, one at the start of 2002 and one in 2005. The cause of these spikes was not thoroughly investigated, but a likely cause is a massive cleanup and refactoring of the database. This is a warning sign that the data, exactly at these points, might contain a lot of noise. In general, the data before 2006 should be considered with more caution than the data afterwards.

@@

Since it was impossible to get a visual overview of the entire dependency graph, its structure was plotted using histograms. Figure~\ref{fig:dephisto-out} is a histogram of the number of dependencies per package. A log-log scale was required to make the plot insightful reflecting the fact that there are many packages with only zero, one or a few dependencies and a few packages with a lot of dependencies. Statistically this means the distribution of the number of dependencies has a fat tail. Likewise, the number of dependers for each package is plotted in figure~\ref{fig:dephisto-in}. This can be interpreted as the distribution of the number of adopters of a given technology. Again, there is a fat tail, even fatter than the one from the number of dependencies. The approximate linearity of the histogram suggests a power-law like distribution of the number of adopters for a given technology. \citet{zheng08} suggest that the structure of package dependency networks is not comparable to known models of social networks and have developed their own model to explain the graph structure.

@@

The program Tulip was used to visualise the graph, the result can be seen in figure~\ref{fig:kde2}. First the graph was laid out using a force based method, this clusters packages that are strongly connected close to each other. Then the graphs where coloured according to their $k$-core measure, this is a measure for the `connectedness' of a package. At this point there was still an unclear mess of lines between the packages, which was resolved by bundling the edges. Edge bundling merges neighbouring lines to a single thicker line and this creates a vein-like structure.

@@

In figure~\ref{fig:kde2} one can see that all the packages depend on \verb|kdelibs|, the large blue dot in the middle. The \verb|kdelibs| package provides a lot of basic functionality, such as a unified set of icons, file open/save dialogues and fewer visible standard components. Almost all the packages in the KDE set require one or more of these components. It should be stressed that there was no manual work involved in the layout of this graph, Tulip was able to determine using only objective, deterministic mathematical methods from graph theory that \verb|kdelibs| plays a central role in the KDE technology.

@@

From the entire list of packages, a few well-known (at least according to the authors) packages were selected. The selection criteria was that the package must not have existed before (around) 2004, because the Gentoo Portage database was still too immature then, and that the package must have gained a considerable number of dependers since its introduction. In table \ref{tbl:results}, we list some the packages along with their parameters. In these cases the adjusted coefficient of determination $\overline{R}^2$ was more than $99\%$.

Special tools where developed to extract the time series of the number of adopters $A_t$ for a given package. This time series was then fit to equation \eqref{eq:bass-model} using Mathematica's \texttt{NonLinearModelFit}. The goodness-of-fit was analysed using an ANOVA table and calculated using the adjusted coefficient of determination $\overline{R}^2$. The parameters were extracted from the fit, and confidence intervals were calculated under assumptions of normality. In table \ref{tbl:results} the relevant parameters are presented with their mean value and a $95\%$ confidence interval. Since normality was assumed, the confidence intervals ignore the $p \ge 0$ and $q \ge 0$ constraints.

@@

The first package we consider is \verb|git|, a modern revision control system that shows an imitator driven adoption. It's growth can be seen in figure \ref{fig:git}, the corresponding statistics are in table \ref{tbl:results}. The package first appeared just before 2005, it had around ten packages depending on it in 2006, twenty in 2008 and is currently used by almost three hundred packages. According to the Bass model, it will continue to grow to approximately 750 users. The innovator inflow is only $0.2\%$ of the potential market per year, so one would expect $0.002\cdot (750 - 300) = 1$ users to adopt \verb|git| out of sheer innovation. Taking the analogy to people, if someone from the $450$ current non-users were to meet a random person from the entire $750$ market, there is a $\frac{300}{750}= 40\%$ chance of meeting a user which can convince him/her to start using \verb|git|. The chance of this happening is the imitator inflow $q = 0.73$. Therefore, the total number of users \verb|git| can expect to gain from imitation this year is $450\cdot 40\%\cdot 0.73 = 131$. Implying that the market is very much imitator driven.

@@

A typical example of innovator driven growth is given by \verb|libmad|. The model is fitted resulting in figure \ref{fig:libmad}. Again, the data is neatly explained by a Bass diffusion process and in particular, the rapid steep growth and the stable user base afterwards. The name of the project is an acronym for ``library for MPEG Audio Decoding'' and the package provides a high quality mp3 decoder for use in multimedia applications. This might also explain the rapid growth of its adoption: multimedia applications can benefit a lot from good quality mp3 support.

@@

Excluding the blue dots from the data results in the Bass model fit from figure \ref{fig:xulrunner}. The fitness increases to $\overline{R}^2 = 99.54\%$ and the parameters have tighter and reasonable confidence intervals. This is strong evidence that the initial adoption of the package is still a Bass diffusion process. To explain the last part, the model should be extended with an obsolescence term.

@@

The manually rejected packages are listed in table \ref{tbl:rejected}. It should be noted that the selection criteria are chosen strictly to ensure that the main hypothesis is tested correctly, with independent open source software projects taken from their introduction and in their whole. When looking at the rejects, it does appear that some of the (\texttt{app-text/asciidoc} packages and many of the \texttt{x11-libs/} packages) still follow the Bass diffusion model.

@@

The package dependency graph contains empirical data to test extensions of the Bass diffusion model - extended with discarders. The Bass model and the present analysis is formulated in terms of absolute number of users, but in most applications only sales figures are available. The amount of sales is the first derivative of the Bass model, hence the model is usually applied in its derivative form \citep{mahajan95}. As a consequence, the model only considers adopters, but does not consider \emph{discarders}. In the \texttt{xulrunner} example, we see the package being discarded from 2011 onwards, providing insights into the discarding mechanism. The next step would be to collect more examples of packages being discarded, and looking at the patterns of demise and perhaps developing a model of discarding to supplement the Bass model of adoption. One model could for example be the inverse of a Bass curve, this makes sense when the market share of the original package is taken over by a new package. The unique feature of dependency graph analysis to give absolute user numbers facilitates this. Such an analysis can also help in predicting the behaviour of certain projects - hence forewarning the particular project stakeholders. As the extent of adoption of an open source project is an indication of how popular the project is and in-turn an indication of the health of the project as assessed by the open source community, we can use this predictive adoption to typically supplement the existing analysis of open source project health \citep{daniel13,amrit10a,piggot13}.

@@

Table~\ref{tbl:pkgmngrs} lists the popular Linux distributions. For each distribution a number of details is given in the columns. The date of its first published version is given. The `package manager' refers to the system used to install and maintain software in a running system. This is relevant for this paper since the package manager must know what packages depend on what other packages so that it installs all the prerequisites. The `source/binary' column specifies whether the packages are installed by compiling the source code or if the distribution provides pre-compiled binaries. The later case is more popular since compiling software is very slow and error prone compared to simply copying pre-compiled files. The last columns list the number of packages in the package database at the time of writing and the average duration between new versions of the package database. Gentoo is unique in that it does not periodically release a new version of the package database, but updates its database as new versions of packages become available.

@@

The Gentoo portage database and the FreeBSD ports databases are ideal candidates. Both have all the good qualities of a source based package database. FreeBSD has the advantage that it has a long history and more packages, Gentoo has the advantage that it has better time resolution. Also since the first author is very familiar with Gentoo's database, we chose the Gentoo portage database as the data source for this article.

>>>>

\title{Diffusion of Innovation in the Open Source Software Ecosystem}
\subtitle{An Empirical Study of the Gentoo Package Database} 
\dedication{}
\author{Remco Bloemen \and Chintan Amrit}
\institute{University of Twente, The Netherlands}

@@

In this paper we make the case that software package dependencies are a form of  adoption of innovation. We then test this on the time-evolution of the Gentoo package dependency graph. We find that the Bass model of innovation diffusion fits the growth of the number of packages depending on a given library with an $\overline{R}^2 > 98\%$. This finding provides open source project users and managers with the information and the possibility to not only identify the innovative open source libraries, but also to predict the the adoption of an open source library over its growth phase.
\end{abstract}

\begin{keywords}{
Open Source Adoption; Innovation Diffusion; Open Source package dependency; Bass diffusion model}
\end{keywords}

@@

In this paper we not only analyse the adoption of open source libraries by other projects, but also use a method to test the diffusion of innovation of a software project - or the lifecycle of a project's adoption. We analyse the adoption of the open source libraries using the theoretical lens of \emph{diffusion of innovation} \citep{rogers03} and test it with a particular formalisation of the theory borrowed from the field of marketing research, namely the \emph{Bass diffusion model} \citep{bass69}. Our central hypothesis is that  \emph{the growth in the number of open source software dependees of a project (its adoption) is a Bass innovation diffusion process.} In particular, we study  software innovation of the libraries in the Gentoo Portage  database from an adoption perspective. We first download the Gentoo Portage database for a time line of one month over a period of around 13 years (from 1 January to around mid 2012). We then calculate the dependencies among the libraries in each version of the database. Then, we calculate the change in the dependency graph of some well known libraries and check if this adoption of the library fits the Bass diffusion model \citep{bass69}.  We then analyse the adoption of 10 randomly selected libraries (we discard and replace those libraries whose adoption trends contain a lot of noise due to various reasons) compared to the Bass diffusion model. In all cases we find that the Bass diffusion model explains the adoption of the libraries (with an $\overline{R}^2 > 98\%$). This finding shows that the adoption of open source software libraries (in the Gentoo Portage database) does not differ from the adoption of other kinds of innovations as predicted by the theory of adoption of innovation, and in particular the Bass diffusion model. This finding, if found generalisable to most large open source libraries, can help in predicting their adoption over their growth phase. This can hence provide guidance to package managers and users of open source libraries in identifying the innovative libraries, as well as predicting their growth adoption and the time of potential saturation and/or decline.

% (Adoption reflects innovation 'better' than changes within packages. - Still cannot find a relevant reference!)
In the next section (section 2)  we describe the background literature of the paper, this is followed by the details of our method of extracting the dependencies of libraries in section 3. In section 4 we analyse the dataset and test the fit of the trends of library adoption with the Bass diffusion model. Section 5 discusses the results and in section 6 we conclude the paper by discussing the limitations of this study and provide some guidance for future work.

@@

\citet{crowston08} have done a comprehensive overview of academical research on open source software development. Of the 184 articles they cite, the vast majority of articles are case studies or surveys, with 4\% of the articles describing the development of empirical instruments and/or measurements. Also, most articles look at the level of a particular group or project, with 4\% looking at the societal level of interacting projects. Crowston et al. (2008) find no articles that develop instruments or measurements on the level of OSS packages \citep{crowston08}, which we aim to do in this paper.

@@

In Figure~\ref{fig:bass} two Bass diffusions are plotted using equation \eqref{eq:bass-direct}, one representing an innovation diffusion with many innovators and a few imitators and one representing a diffusion with many imitators and a few innovators. When there are more imitators it takes a while for the innovation to be adopted since the majority of the potential users are waiting for someone else to try it first.

@@

The interpretation of the variables and parameters and their dimensions is presented in table~\ref{tbl:bass-model}. When applying the formula, one should note that it is non-linear, so ordinary linear regression can not be used. Instead one can use non-linear least squares regression, while noting the correct number of the degrees of freedom. This can be done using existing mathematical/statistical packages. In this paper we used the \verb|NonlinearModelFit| procedure in Mathematica. This is a variant of the \emph{Nonlinear least squares} method that has been shown to be more optimal than an \emph{Ordinary least squares} fit and in some cases better than a \emph{Maximum likelihood estimation}  \citep{meade06}.

@@

By design the database can work with complicated dependency relations, such as ``package \verb|amarok| requires package \verb|phonon-kde|, minimum version \verb|4.3|, but only when feature \verb|player| is required''. This would be coded as \verb|player?| \verb|(| \verb|>=kde-base/phonon-kde-4.3| \verb|)|. An example of the run time dependencies for the Amarok music player is given in Figure~\ref{fig:amarokruntime}. The figure includes more complex rules such as ``either package X or package Y is required'', `` package X is required to contain feature Y'', etcetera.

@@

In Figure~\ref{fig:pkgsgrowth} the number of packages in the database is plotted over time. One can see how the database started in 2001, underwent a period of rapid growth between 2002---2006 and settled into calm a linear growth from 2006 onwards. In Figure~\ref{fig:pkgsgrowthdelta} the rate of introducing and removing packages is plotted over time. This shows two spikes, one at the start of 2002 and one in 2005. The cause of these spikes was not thoroughly investigated, but a likely cause is a massive cleanup and refactoring of the database. This is a warning sign that the data, exactly at these points, might contain a lot of noise. In general, the data before 2006 should be considered with more caution than the data afterwards.

@@

Since it was impossible to get a visual overview of the entire dependency graph, its structure was plotted using histograms. Figure~\ref{fig:dephisto-out} is a histogram of the number of dependencies per package. A log-log scale was required to make the plot insightful reflecting the fact that there are many packages with only zero, one or a few dependencies and a few packages with a lot of dependencies. Statistically this means the distribution of the number of dependencies has a fat tail. Likewise, the number of dependers for each package is plotted in Figure~\ref{fig:dephisto-in}. This can be interpreted as the distribution of the number of adopters of a given technology. Again, there is a fat tail, even fatter than the one from the number of dependencies. The approximate linearity of the histogram suggests a power-law like distribution of the number of adopters for a given technology. \citet{zheng08} suggest that the structure of package dependency networks is not comparable to known models of social networks and have developed their own model to explain the graph structure.

@@

The program Tulip was used to visualise the graph, the result can be seen in Figure~\ref{fig:kde2}. First the graph was laid out using a force based method, this clusters packages that are strongly connected close to each other. Then the graphs where coloured according to their $k$-core measure, this is a measure for the `connectedness' of a package. At this point there was still an unclear mess of lines between the packages, which was resolved by bundling the edges. Edge bundling merges neighbouring lines to a single thicker line and this creates a vein-like structure.

@@

In Figure~\ref{fig:kde2} one can see that all the packages depend on \verb|kdelibs|, the large blue dot in the middle. The \verb|kdelibs| package provides a lot of basic functionality, such as a unified set of icons, file open/save dialogues and fewer visible standard components. Almost all the packages in the KDE set require one or more of these components. It should be stressed that there was no manual work involved in the layout of this graph, Tulip was able to determine using only objective, deterministic mathematical methods from graph theory that \verb|kdelibs| plays a central role in the KDE technology.

@@

From the entire list of packages, a few well-known (at least according to the authors) packages were selected. The selection criteria was that the package must not have existed before (around) 2004, because the Gentoo Portage database was still too immature then, and that the package must have gained a considerable number of dependers since its introduction. In Table \ref{tbl:results}, we list some the packages along with their parameters. In these cases the adjusted coefficient of determination $\overline{R}^2$ was more than $99\%$.

Special tools where developed to extract the time series of the number of adopters $A_t$ for a given package. This time series was then fit to equation \eqref{eq:bass-model} using Mathematica's \texttt{NonLinearModelFit}. The goodness-of-fit was analysed using an ANOVA table and calculated using the adjusted coefficient of determination $\overline{R}^2$. The parameters were extracted from the fit, and confidence intervals were calculated under assumptions of normality. In Table \ref{tbl:results} the relevant parameters are presented with their mean value and a $95\%$ confidence interval. Since normality was assumed, the confidence intervals ignore the $p \ge 0$ and $q \ge 0$ constraints.

@@

The first package we consider is \verb|git|, a modern revision control system that shows an imitator driven adoption. It's growth can be seen in Figure \ref{fig:git}, the corresponding statistics are in Table \ref{tbl:results}. The package first appeared just before 2005, it had around ten packages depending on it in 2006, twenty in 2008 and is currently used by almost three hundred packages. According to the Bass model, it will continue to grow to approximately 750 users. The innovator inflow is only $0.2\%$ of the potential market per year, so one would expect $0.002\cdot (750 - 300) = 1$ users to adopt \verb|git| out of sheer innovation. Taking the analogy to people, if someone from the $450$ current non-users were to meet a random person from the entire $750$ market, there is a $\frac{300}{750}= 40\%$ chance of meeting a user which can convince him/her to start using \verb|git|. The chance of this happening is the imitator inflow $q = 0.73$. Therefore, the total number of users \verb|git| can expect to gain from imitation this year is $450\cdot 40\%\cdot 0.73 = 131$. Implying that the market is very much imitator driven.

@@

A typical example of innovator driven growth is given by \verb|libmad|. The model is fitted resulting in Figure \ref{fig:libmad}. Again, the data is neatly explained by a Bass diffusion process and in particular, the rapid steep growth and the stable user base afterwards. The name of the project is an acronym for ``library for MPEG Audio Decoding'' and the package provides a high quality mp3 decoder for use in multimedia applications. This might also explain the rapid growth of its adoption: multimedia applications can benefit a lot from good quality mp3 support.

@@

Excluding the blue dots from the data results in the Bass model fit from Figure \ref{fig:xulrunner}. The fitness increases to $\overline{R}^2 = 99.54\%$ and the parameters have tighter and reasonable confidence intervals. This is strong evidence that the initial adoption of the package is still a Bass diffusion process. To explain the last part, the model should be extended with an obsolescence term.

@@

The manually rejected packages are listed in Table \ref{tbl:rejected}. It should be noted that the selection criteria are chosen strictly to ensure that the main hypothesis is tested correctly, with independent open source software projects taken from their introduction and in their whole. When looking at the rejects, it does appear that some of the (\texttt{app-text/asciidoc} packages and many of the \texttt{x11-libs/} packages) still follow the Bass diffusion model.

@@

The package dependency graph contains empirical data to test extensions of the Bass diffusion model - extended with discarders. The Bass model and the present analysis is formulated in terms of absolute number of users, but in most applications only sales figures are available. The amount of sales is the first derivative of the Bass model, hence the model is usually applied in its derivative form \citep{mahajan95}. As a consequence, the model only considers adopters, but does not consider \emph{discarders}. In the \texttt{xulrunner} example, we see the package being discarded from 2011 onwards, providing insights into the discarding mechanism. The next step would be to collect more examples of packages being discarded, and looking at the patterns of demise and perhaps developing a model of discarding to supplement the Bass model of adoption. One model could for example be the inverse of a Bass curve, this makes sense when the market share of the original package is taken over by a new package.

The unique feature of dependency graph analysis to give absolute user numbers facilitates this. Such an analysis can also help in predicting the behaviour of certain projects - hence forewarning the particular project stakeholders. As the extent of adoption of an open source project is an indication of how popular the project is and in-turn an indication of the health of the project as assessed by the open source community, we can use this predictive adoption to typically supplement the existing analysis of open source project evolution \citep{deshpande08,amrit10a, godfrey00, amrit10b} as well as open source project health \citep{daniel13,piggot13,subramaniam09}. A related implication of our finding is that it provides open source project user and managers with the information to also identify the innovative open source libraries. This is especially true when the open source project is being adopted in the early growth stage.

@@

Table~\ref{tbl:pkgmngrs} lists the popular Linux distributions. For each distribution a number of details is given in the columns. The date of its first published version is given. The `package manager' refers to the system used to install and maintain software in a running system. This is relevant for this paper since the package manager must know what packages depend on what other packages so that it installs all the prerequisites. The `source/binary' column specifies whether the packages are installed by compiling the source code or if the distribution provides pre-compiled binaries. The latter case is more popular since compiling software is very slow and error prone compared to simply copying pre-compiled files. The last two columns list the number of packages in the package database at the time of writing and the average duration between new versions of the package database. Gentoo is unique in that it does not periodically release a new version of the package database, but updates its database as new versions of packages become available.

@@

The Gentoo portage database and the FreeBSD ports databases are ideal candidates, as both are source distributions. Both have all the good qualities of a source based package database. FreeBSD has the advantage that it has a long history and more packages, Gentoo has the advantage that it has a better time resolution. Also since the first author is very familiar with Gentoo's database, we chose the Gentoo portage database as the data source for this article.
